Income Disparity and Its Influence on Higher Education Affordability and Access

### 2024-02-19 starter_folder-main 4/paper/college_tuiton_income_inequality.qmd
added variables

@@

R was the language and environment used for the bulk of this reproduction, alongside

## Features

@@

## Limitations to reproducing

>>>>

This paper reproduces several areas of the original paper which is orginaially from the American Economic Association, utilizing the same datasets in the process. These datasets are unfortuntely not publicly accessible, and are hidden behind a sign in wall.

@@

R was the language and environment used for this reproduction, alongside different packages such as ggplot2, readxl, dplyr, tidyr, knitr packages.

## Variables

Variables within the datasets have been constructed to reflect the nuances of income distribution across different socioeconomic groups and the corresponding tuition rates at a range of institutions, from community colleges to elite universities.

We have listed the different variables used below and how they correspond:

1. Net Prices Data:

public_college_netprice: Data on the net prices of public colleges.

private_college_netprice: Data on the net prices of private colleges.

2. Grants Data:

public_college_grants: Data on total grants by income and year for public four-year colleges.

private_college_grants: Similar data for private four-year colleges.

3. Funding and Revenue Data:

state_funding: average state and local funding for higher education per 1000$ of personal income

public_rev_subs and private_rev_subs: detailing subsidies and revenues for public and private four-year institutions, with columns for institution type, year, net tuition revenue, subsidy, and education-related expenditures.

4. College Tuition and Fees Data:

Public Sticker: This is the advertised or sticker price for attending a public college or university. It includes tuition, fees, room and board, and other expenses before any financial aid, scholarships, or grants are applied. This price is typically higher than what most students will pay, especially after financial aid is taken into account.

Public Net: The net price for attending a public college or university. This is the amount that students and their families actually pay on average after subtracting scholarships, grants, and financial aid from the public sticker price. The net price provides a more accurate representation of the cost to students and families after accounting for available financial assistance.

Private Sticker: Similar to the public sticker price, this is the advertised price of attending a private college or university. It tends to be higher than the public sticker price, reflecting the total cost of tuition, fees, room and board, and other expenses at a private institution before any financial aid is applied.

Private Net: The net price for attending a private college or university. It's the actual cost to students and their families after deducting scholarships, grants, and financial aid from the private sticker price. Since private colleges often offer significant financial aid packages, the private net price can be considerably lower than the sticker price, though it generally remains higher on average than the public net price.

5. Income Distribution Data

Log Household Income: This refers to the log-transformed household income. In economic data analysis, income values are often log-transformed for several reasons such as to reduce skewness and helps linearize multiplicative relationships, making them easier to model and interpret

Density (EMG_ML_PDF): EMG_ML_PDF" stands for the Estimated Maximum Likelihood Probability Density Function. This term combines several concepts such as Estimated Maximum Likelihood (EML) which is a statistical method used to estimate the parameters of a probability distribution that make the observed data most probable. It's a common approach in econometrics and statistics for fitting models to data and Probability Density Function (PDF) which describes the likelihood of a random variable to take on a given value. In practical terms, it shows how the values of the random variable are distributed across different outcomes. This density variable shows how the probability distribution of incomes (after log transformation) changes across different income levels for the years analyzed.

@@

## Limitations to our reproduction
